---
title: "iris-dataset-analysis"
author: "Adrian M"
date: "2023-10-14"
output: github_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# install.packages("tidyverse")
# install.packages("ggplot2")
# library(tidyverse)
library(ggplot2)
```

Copying iris data set to variable 'df'

```{r}
df <- iris
```

A quick view of the dataset

```{r}
head(df)
```
Trying univariate analysis and visualize this using qplot

?qplot

Using qplot to show a plot. Check the number of records per Species

```{r}
qplot(Species, 
      fill = Species, #to identify the color per species
      data = df)
```

qplot automatically decides which plot to use based on supplied variables. 
50 records per species.

Trying qplot if it will show a scatterplot

```{r}
qplot(Petal.Length, 
      Petal.Width, 
      data = df)
```

Add some colors by Species

```{r}
qplot(Petal.Length, 
      Petal.Width, 
      color = Species,
      data = df)
```

Looks good but documentation says qplot is deprecated (no longer useful or better options exist)
So, going to try and explore ggplot instead

?ggplot

```{r}
#add the base layer
ggplot(data = df, 
       mapping = aes(x = Petal.Length, 
                     y = Petal.Width))
```

It's blank. Of course since the `aes()` or aesthetic is only present here. I need to specify which type of plot to use (unlike qplot?)

```{r}
ggplot(data = df, 
       mapping = aes(x = Petal.Length, 
                     y = Petal.Width)) +
  geom_point() #scatterplot layer
```

Pros: more control over what you want to display versus qplot which does it automatically.
Also, it says we can drop the specifications for the `data=` and `mapping=` arguments as long as we supply the correct arguments in the correct order. Nice. 
Add some colors

```{r}
ggplot(df,
       aes(x = Petal.Length,
           y = Petal.Width,
           col = Species)) + #color that helps identify by species
  geom_point()
```

Add a trend line since it's a scatterplot, to see the correlation better

```{r}
ggplot(df, 
       aes(x = Petal.Length,
           y = Petal.Width,
           col = Species)) +
  geom_point() + 
  geom_smooth(method = lm) #add the trend line layer
```

All three species appears to have a positive correlation for Petal length & width shown by the trend line going upwards.

I can drop the `x=` and `y=` specifications too for the `aes()` method.

?aes()

Also, ggplot uses layers so it can overlap multiple plots, from one dataset or more(!). I'll to move the `aes()` inside the `geom_point` layer and try some parameters

```{r}
ggplot(df) +
  geom_point(aes(Petal.Length,
                 Petal.Width,
                 col = Species, #color by species 
                 shape = Species, #shape by species
                 size = 3, #size of datapoints
                 stroke = 1, #thickness of borders of datapoints, specific to geom_point
                 alpha = 0.7)) #transparency/opacity     
```

Using facet to separate

?facet_wrap

```{r}
ggplot(df,
       aes(Petal.Length,
           Petal.Width,
           col = Species)) +
  geom_point() +
  facet_wrap(Species ~.) + #separates by Species into own grids/plots
  theme(legend.position = "bottom") #move the legend in the bottom
```

Try to overlap two layers, petal and sepal 

```{r}
ggplot(df) +
  geom_point(aes(Petal.Length,
             Petal.Width,
             col = Species)) +
  geom_point(aes(Sepal.Length,
             Sepal.Width,
             col = Species))
```

I can't distinguish which is which. Maybe try to change the color per variable

colors()

```{r}
ggplot(df) +
  geom_point(aes(Petal.Length,
                 Petal.Width,
                 shape = Species), 
             col = "seagreen") +        #put the col= argument outside aes()
  geom_point(aes(Sepal.Length,
                 Sepal.Width,
                 shape = Species),
             col = "royalblue")
```

This look too busy and unusable but I can move forward and explore ggplot further.

Try to see some relation between the measurements (petal, sepal, length and width) but this time, show the plots side by side using cowplot package

install.packages("cowplot")

```{r}
library(cowplot)
```

```{r}
gg1 <- ggplot(df, aes(col = Species)) +
  geom_point(aes(Petal.Length,
                 Petal.Width), 
             size = 2, 
             alpha = 0.7,
             show.legend = FALSE)

gg2 <- ggplot(df, aes(col = Species)) +
  geom_point(aes(Sepal.Length,
                 Sepal.Width), 
             size = 2, 
             alpha = 0.7) +
  theme(legend.position = "top",
        legend.title = element_blank())

plot_grid(gg1, gg2, ncol = 2) #from cowplot
```

Looks crowded but I can see here that Petal Length & Width have positive correlation while Sepal length & width have positive correlation only with setosa 

OK I'm going to try histogram plots. Some basic plots to see the distribution of all the measurements

```{r}
#I made a function since this will repeat 4 times (petal & sepal, width & length)
create_histogram <- function(data, column_name, num_bins = 25) #create the function along with the accepted parameters
  ggplot(data) +
  geom_histogram(aes_string(column_name), 
                 bins = num_bins,
                 alpha = 0.7) + #flexibility in number of histogram bins
  theme_minimal() #makes all plots consistent

#Running the function, supplying the appropriate parameters, then storing each plot to a corresponding variable
hist_petal_length <- create_histogram(df, "Petal.Length")
hist_petal_width <- create_histogram(df, "Petal.Width")
hist_sepal_length <- create_histogram(df, "Sepal.Length")
hist_sepal_width <- create_histogram(df, "Sepal.Width")

#Now plot the 4 in a 2x2 grid using plot_grid from cowplot
plot_grid(hist_petal_length, 
          hist_petal_width, 
          hist_sepal_length, 
          hist_sepal_width, 
          ncol = 2, 
          nrow = 2)

```
The petal length & width appears to have a gap in a certain section of the distribution. The group on the left could be the setosa, similar to the gaps on the previous scatterplots. 
The sepal length is spread across the measurements but the width looks to have a central tendency towards "3"

I'll take a closer look at petals using overlapping density plot

```{r}
ggplot(df) +
  geom_histogram(aes(x = Petal.Length,
                   y = after_stat(density),
                   fill = Species),
               bins = 30,
               alpha = 0.5,
               position = "identity") + 
  geom_density(aes(x = Petal.Length, 
                   col = Species),
               size = 1.2)
```
Identifying the species by colors confirms that the gap is because of the iris specie setosa

Can we try using box plots to explore this dataset more?

?geom_boxplot

```{r}
#make another function for the boxplot
create_boxplot <- function(data, column_name, fillBy = "Species")
  ggplot(data,
       aes_string(x = fillBy,     
           y = column_name, 
           fill = fillBy)) + 
  geom_boxplot() + 
  theme_minimal() +
  theme(legend.position = "none") +
  coord_flip() #swap the axes to make the display horizontal

gg_boxp_petalL <- create_boxplot(df, "Petal.Length")
gg_boxp_petalW <- create_boxplot(df, "Petal.Width")
gg_boxp_SepalL <- create_boxplot(df, "Sepal.Length")
gg_boxp_SepalW <- create_boxplot(df, "Sepal.Width")

plot_grid(gg_boxp_petalL,
          gg_boxp_petalW,
          gg_boxp_SepalL,
          gg_boxp_SepalW,
          ncol = 2,
          nrow = 2)
```
It looks normally distributed but there are possible outliers. Taking at closer look at Sepal width

```{r}
ggplot(df,
       aes(
         x = Species,
         y = Sepal.Width,
         fill = Species)) +
  geom_boxplot() +
  geom_jitter(position = position_jitter(width = 0.2),
              alpha = 0.7) +
  theme_minimal() +
  coord_flip()
```
Some data points are clearly outside lines extending outwards the box but I cannot conclude that these are erroneous data. Since I cannot gather more information about these "outlier" measurements for this dataset, I will assume they are correct data for now and for the rest of the box plots.


