---
title: "mtcars-data-anaysis"
author: "Adrian Manriza"
date: "2023-10-16"
output: github_document
editor_options: 
  chunk_output_type: inline
---

# Introduction

Motor Trend Car Road Tests

The data was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973–74 models).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

pacman::p_load(tidyverse, ggplot2)
```

Looking at the dataset

```{r}
head(mtcars)
```

```{r}
summary(mtcars)
```

First, I want to try cluster analysis.

Clustering, meaning to group (or cluster) similar items together based on their characteristics. 
The goal is to partition a dataset into groups, or clusters, where items in the same group are more similar to each other than to those in other groups.

Preparing data for hierarchical cluster

```{r}
dfcars <- mtcars %>%
  rownames_to_column() %>% #add a column name instead of a rowname to car names
  as_tibble() %>% #save as tibble
  select(car = rowname, mpg:hp, wt, qsec, gear, carb) %>% #choosing specific columns
  mutate_at(vars(-car), scale) %>% #make all <dlb> type columns have same scale using z-score
  print()
```

Calculating

```{r}
hc <- dfcars %>%
  dist %>% #distance
  hclust #agglomerative, analyze single then group
```

Plotting dendogram

```{r}
invisible(dev.hold()) #keep the graphics device open
#plot the car names
hc %>% plot(labels = dfcars$car, 
           cex = 0.8, #label size
           hang = -1) #bottom alignment

#draw boxes
hc %>% rect.hclust(k = 5, #k = number of groups
                   border = 2:6) #border color, colors 2-6 from color palette
invisible(dev.flush()) #close the graphics device
```

